---
title: "Chicago Bikeshare Analysis - Divvy Bike Usage in Chicago's Transit Deserts"
author: "Markisha (Misha) Berrien"
date: "2/3/2019"
output:
  html_document:
    code_folding: hide
---

```{r include=FALSE}
# Set-up Environment 
#knitr::opts_chunk$set(include = FALSE)
library(tidyverse)
library(lubridate)
library(geosphere)
library(ggrepel)
library(forcats)
library(devtools)
library(extrafont)
library(ggmap)
# devtools::install_github('cttobin/ggthemr') # This sets a theme for the entire session

#register_google(key = "AIzaSyDu8dhum7j58jpfpQq6aE2jcZLgku9wzzI")

```

```{r eval=FALSE, include=FALSE}
# Save large dataset
divvy_trips <- 
  read.csv("Divvy_Trips.csv")

saveRDS(divvy_trips, file = "divvy_trips.Rda")
```


```{r warning=FALSE, message=FALSE, error=FALSE}
# Read in datasets
bus_ridership <- 
  read.csv("/Users/Misha/Documents/04-coding-projects/chicago-bikeshare-analysis/code-data/cta_bus_ridership_dailytots.csv")

l_ridership <- 
  read.csv("/Users/Misha/Documents/04-coding-projects/chicago-bikeshare-analysis/code-data/cta_l_ridership_dailytots.csv")

divvy <- 
  readRDS("/Users/Misha/Documents/04-coding-projects/chicago-bikeshare-analysis/code-data/divvy_trips.Rda")

chi_ca <- 
  read_csv("/Users/Misha/Documents/04-coding-projects/chicago-bikeshare-analysis/code-data/chi_community_areas.csv")

divvy_stations <- 
  read_csv("/Users/Misha/Documents/04-coding-projects/chicago-bikeshare-analysis/code-data/Divvy_Bicycle_Stations_In_Service.csv")

census_2010 <- 
  read_csv("/Users/Misha/Documents/04-coding-projects/chicago-bikeshare-analysis/code-data/chi_pop_ca.csv")

l_stop <- 
  read_csv("/Users/Misha/Documents/04-coding-projects/chicago-bikeshare-analysis/code-data/CTA_L_STOPS.csv")

cta_bus <- 
  read_csv("/Users/Misha/Documents/04-coding-projects/chicago-bikeshare-analysis/code-data/CTA_BusStops.csv")

  # Take 5% Sample of divvy dataset
divvy_sample <- 
  sample_frac(divvy, size = .05)

# Add Community Area and population to divvy dataset
divvy_census <-
  divvy_sample %>%
  left_join(chi_ca, by = "Community.Areas")

divvy_census <- 
  divvy_census %>%
  left_join(census_2010, by = "Community.Areas")
```



```{r warning=FALSE, message=FALSE, error=FALSE}
# Create color palette
color_dis <- c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", "#fdbf6f")


bikeshare_style <- function() {
  theme(
    axis.line = element_line(size = 0.05),
    text = element_text(family = "SourceSerifPro-Regular", face = "plain", 
                              colour = "black", size = 10, 
                              hjust = 0.5, vjust = 1, angle = 0, 
                              lineheight = 0.9), 
    plot.subtitle = element_text(family = "SourceSerifPro-Regular",
                               size = 8), 
    legend.background = element_rect(fill = "#DBDBDB", color = "black", size = 0.1),
    
    axis.text = element_text(family="SourceSerifPro-Regular",
                           size=8,
                           color="#222222"),
    axis.text.x = element_text(vjust = 1), 
    axis.ticks = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "black", size = 0.1),
    panel.background = element_rect(fill = "#F2F2F2", colour = NA),
    
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size  = 22,  hjust = 0),
    plot.background = element_rect(colour = "white", fill = "#F2F2F2", size = 1),
    plot.margin = unit(c(2, 2, 1, 1), "lines")
  )
  }
```


On average, the South Side accounts for some of the shortest divvy bike trips in the city, while the West Side and Far Southeast Side account for some of the longest. 
```{r warning=FALSE, message=FALSE, error=FALSE}
#parse date columns
divvy_date <- 
  divvy_census %>%
  mutate(dateatime = START.TIME) %>%
  separate(dateatime, into = c("date", "time", "ampm"), sep = " ") %>%
  separate(date, into = c("month", "day", "year", sep = " "))

# Create variable "distance_hav" which shows distance bikeride traveled (meters)
divvy_distance <- 
  divvy_date %>%
  filter(!is.na(Community.Areas)) %>%
  mutate(distance_meters = 
           distHaversine(matrix(
             c(FROM.LONGITUDE, FROM.LATITUDE), ncol = 2),
                                   matrix(
                                     c(TO.LONGITUDE, TO.LATITUDE), ncol = 2)))

divvy_distance %>%
  filter(!is.na(distance_meters)) %>%
  filter(year == 2018) %>%
  group_by(Chicago.side) %>%
  summarise(avg_distance_meters = mean(distance_meters)) %>%
  ggplot(aes(x = reorder(Chicago.side, avg_distance_meters), y = avg_distance_meters, 
             fill = Chicago.side), fill = "#2c7bb6") +
  geom_point() +
  theme(legend.position = "none") +
  labs(title = "The South Side accounts for the shortest average Divvy \ntrips within the City of Chicago",
    subtitle = "Average distance traveled by divvy trips in 2018",
    caption = "Chicago Data Portal Divvy Trip Data",
    x = "Chicago 'Sides'",
    y = "Average Divvy Trip Distance (Meters)") +
  coord_flip() +
  bikeshare_style()
  #scale_y_continuous(breaks = c(1, 2, 3))

```

Although the South Side accounts for some of the shortest bike trips in terms of distance, this area accounts for some of the longest trip duration in minutes. A divvy membership includes 30 minutes of "free riding time" from bike check-out to return. In the majority of the Chicago areas, the average ride duration hovers just around the 30-minute mark, but the South, Southwest and West Side do not follow this trend. 
```{r warning=FALSE, message=FALSE, error=FALSE}
divvy_date %>%
  filter(!is.na(Community.Areas)) %>%
  filter(year == 2014 | year == 2015 | year == 2016 | year == 2017) %>%
  filter(!TRIP.DURATION > 1440) %>% # Filter for bikes classified as stolen
  group_by(Chicago.side) %>%
  summarise(average_time = mean(TRIP.DURATION)/60) %>%
  ggplot() +
  geom_col(aes(x = reorder(Chicago.side, average_time), 
      y = average_time, 
      fill = as.factor(Chicago.side)), fill = "#2c7bb6", 
    position = "dodge") +
  coord_flip() +
  labs(title = "Divvy bike trip duration is greater in Chicago's \nNorth and South Suburbs",
    subtitle = "Average Divvy bike trip length from 2014 - 2017",
    caption = "Source: Chicago Data Portal Divvy Trip Data",
    x = "Chicago 'Sides",
    y = "Average Trip Time (In Minutes)") +
  bikeshare_style() +
  scale_y_continuous(expand=c(0,0), breaks = c(0, 2, 4, 6, 8, 10, 12, 14))
```

Upon closer inspection of the three "Chicago Sides" with higher than average rider time, you can see five Community areas that are responsible for increasing the average trip duration time for the South, Southwest and West side. 
```{r warning=FALSE, message=FALSE, error=FALSE}
divvy_census %>%
  filter(!is.na(Community.Areas)) %>%
  filter(!TRIP.DURATION > 1440) %>% # Filter for bikes classified as stolen
  filter(Chicago.side == c("South Side", "Southwest side", "West Side")) %>%
  group_by(Community.Area.Name, Chicago.side) %>%
  summarise(average_time = mean(TRIP.DURATION)/60) %>%
  ggplot(aes(x = average_time, 
             y = reorder(factor(Community.Area.Name), average_time))) + 
  geom_segment(aes(yend = Community.Area.Name, xend = 0)) +
  geom_point() +
  labs(title = "Five community areas in South, Southwest and \nWest Chicago account for above average trip time",
    subtitle = "Average divvy bike trip length from 2013 - 2018 in South, \nSouthwest and West Side Chicago Commuity Areas",
    caption = "Chicago Data Portal Divvy Trip Data",
    x = "Average Trip Time by Community Area (Minutes)",
    y = " ") +
  labs(fill = "Chicago 'Sides'") +
  bikeshare_style() +
  scale_x_continuous(expand = c(0,0))
```


This graph compares divvy ride time duration changes between 2014 and 2017. This shows a strong linkage between number of bike stations in an area and trip duration. All areas show a clear decline in the average trip duration as the number of stations increase. 
```{r warning=FALSE, message=FALSE, error=FALSE}
divvy_stations_year <- 
  divvy_date %>%
  filter(!is.na(Community.Areas)) %>%
  group_by(Chicago.side, year) %>%
  distinct(FROM.STATION.ID) %>%
  summarise(n_stations = n()) 

divvy_date %>%
  filter(!is.na(Community.Areas)) %>%
  filter(year == c(2014, 2017)) %>%
  group_by(Chicago.side, year) %>%
  summarise(average_trip_time = mean(TRIP.DURATION)) %>%
  left_join(divvy_stations_year, by = c("Chicago.side", "year")) %>%
  arrange(Chicago.side, year) %>%
  ggplot(aes(x = year, y = average_trip_time, group = Chicago.side, 
             size = n_stations, 
             color = as.factor(Chicago.side))) +
  geom_point(alpha = 0.6) +
  geom_line(size = 1) +
  labs(title = "Average ride time reduces as number of stations increase",
    subtitle = "Average trip time comparios across all eight 'Chicago Sides' in 2014 and 2017",
    caption="Chicago Data Portal Divvy Trip Data",
    x="Year",
    y="Divvy Trip Time (In Seconds)",
    color = "Chicago 'Side'",
    size = "Number of Stations") +
  bikeshare_style()

```

Share of the city's population does not directly correlate with share of the city's divvy ridership. for example, The South Side makes up less than 10% of the city's population, but accounts for over 40% of the rides in 2018. 
```{r warning=FALSE, message=FALSE, error=FALSE}
div_area <- 
  divvy_date %>%
  filter(!is.na(Community.Areas)) %>%
  filter(year == 2014 | year == 2015 | year == 2016 | year == 2017) %>%
  group_by(Community.Areas) %>%
  summarise(pop = unique(Total.Population), 
            n_rides = n()) %>%
  left_join(chi_ca, by = "Community.Areas") %>%
  group_by(Chicago.side) %>%
  summarise(sum_pop = sum(pop), 
            sum_rides = sum(n_rides)) %>%
  mutate(perc_rides = (sum_rides / sum(sum_rides) * 100), 
            perc_pop = (sum_pop / sum(sum_pop)) * 100)

div_area %>%
  ggplot(aes(x = perc_pop, y = perc_rides, color = Chicago.side)) +
  geom_point(size = 1) +
  theme(legend.position = "none") +
  geom_label_repel(aes(label=Chicago.side)) +
  labs(title="The South Side sccounts for less than 10% of the \npopulation, but accounts for over 40% of rides",
    subtitle="Divvy Ridership from 2014 - 2017",
    caption="Chicago Data Portal Divvy Trip Data",
    x="Chicago Population (by percentage)",
    y="Divvy Bike Rideshare (by percentage)") +
  bikeshare_style() +
  scale_x_continuous(expand=c(0,0), breaks = c(0, 5, 10, 15, 20, 25, 30))
```

Map of bus and train station locations around the city of chicago
```{r warning=FALSE, message=FALSE, error=FALSE}
# Clean l_stop data
#Seperate location variable into latitude and longitude variable
l_stop <- 
  l_stop %>%
  separate(Location, into = c("lat", "lon"), sep = ",") %>%
  mutate(lat = str_remove(lat, "\\(")) %>%
  mutate(lon = str_remove(lon, "\\)"))

# Rename lattitude and longitude lines
cta_bus <- 
  cta_bus %>%
  rename("lon" = "POINT_X") %>%
  rename("lat" = "POINT_Y")

# grab a center/zoom map and compute its bounding box
chi_map <- ggmap(get_googlemap(center = c(-87.682571,41.888605),
                    zoom = 12, scale = 2,
                    maptype ='terrain',
                    color = 'color'))

# Map of bus stops
chi_map + geom_point(aes(x = lon, y = lat,  colour = "red"), 
                     data = cta_bus, size = 0.5) + 
  theme(legend.position = "bottom")
```

